Towards Correct Legion

### 2017-08-26 correctnesshpc/paper.tex
change tone to be more formal

\begin{abstract}
Legion is a programming model that uses logical regions of memory to reason about 
locality and concurrency. We formalize the Legion semantics in a mechanized
proof assistant, Coq, and show how this enables one to reason formally and
prove theorems about the semantics. We also discuss how to extend this work to
prove that serial semantics are preserved and how given enough time, one could
extend the effort to prove an implementation is correct.
\end{abstract}

@@

\section{Intro}
High performance computing, as the name implies, has long focused on
performance. Expensive machines are purchased to run increasingly complex
simulations, and every optimization can save huge amounts of money, both due to
the initial cost of the machine and the cost of power to run it.

That said, what good is a fast computation if we aren't certain that the
values it computes are correct? Presumably, these computations inform important
decisions in areas ranging from climate to nuclear safety. We should strive to
be as confident as possible that we aren't basing these decisions on buggy
code.

The strongest guarantees of correctness come from the area of formal
specifications and proofs. A specification is a description of how a program
should behave, while proofs are used to show that an implementation conforms to
a specification. \emph{Formal} specifications and proofs are defined as
specifications and proofs that have been embedded in a machine-checked formal
logic \cite{bertot2013interactive, nipkow2002isabelle, pfenning1999system,
jackson1994nuprl}. This process of formalizing specifications and proofs in
machine checked logics has been shown to catch common errors in previously
informal proofs, and therefore is often considered to provide the highest
assurance of correctness currently possible \cite{clarke1996formal,
yang2011finding}.

In addition, recent work has shown that modular, composable formal
specifications enable formal correctness proofs for large software systems that
were previously unattainable \cite{weng2016deepspec}. By integrating a formal
specification tightly with an implementation, one can achieve the highest
levels of confidence in code currently possible \cite{weng2016deepspec}. For
example, the Compcert project implements an optimizing C compiler and proves it
correct. The success of Compcert in removing compiler bugs has been well
documented. We will discuss later how Compcert could be used in building
verifiable HPC systems.

In HPC, programming models have become an increasingly important way to ensure
that code stays maintainable, performant, and portable \cite{diaz2012survey}.
We claim that these programming models provide a good target for formalization:
any correctness arguments can be leveraged by every program using the model. It
is exactly this kind of modularization of proofs that has been so successful in
proving programs correct \cite{weng2016deepspec}.

For this paper, we take a recent programming model, Legion, and formalize its
semantics and type system. We then prove a few supporting lemmas as examples of
the kinds of things one can prove in an HPC settings. Finally, we discuss the
some of the possible paths towards a full, formally correct implementation.
This is no small task, as we discuss later, and the ways forward are numerous
and difficult. But if we believe that HPC informs important decisions, we must
rise to the challenge.

@@

Legion is a programming model for high performance computing that enables
reasoning about locality and concurrency via logical regions of memory. It has
the attractive property that it is guaranteed to preserve serial semantics for
memory operations within logical regions. This allows programmers to avoid the
burden of reasoning about the non-determinism that arises from concurrent
computation. For this paper, we'll be formalizating work on from Triecher et.
al \cite{treichler2013language}. While there has been much work extending
Legion since then, the core semantics has stayed relatively stable
\cite{slaughter2015regent, treichler2016dependent}.

By partitioning regions into \emph{subregions}, and tracking \emph{privileges},
Legion is able to ensure that certain code won't interfere or alias, and can
therefore be safely, at least with respect to those memory regions, run in
parallel. This theory has been put into practice to impressive effect, and has
been shown to scale both within a node and between \cite{bauer2012legion,
treichler2014realm}. That said, there are still subtle bugs that occasionally
show up in the implementation. One goal of this paper is to move towards
formally and systematically removing \emph{all} bugs from the programming
model, giving everyone who uses is far more certainty that their code is
correct.

To move towards this goal, we turn to mechanized proof assistants.
Improvements to proof assistants over recent decades have made them a powerful
tool for implementing and reasoning about code. From provably correct compilers
to provably correct operating systems, proof assistants enable extremely high
certainty in correctness even for large scale software projects.

Coq, in particular, has proven itself to be a particularly potent dependently
typed proof assistant. At its core is the Calculus of Inductive Constructions,
a small dependently typed programming language and logic, capable of powerful
abstractions for formal proofs \cite{bertot2013interactive}. In the last 10
years great strides have been made in using it to both build systems and prove
useful properties about them \cite{leroy2012compcert, anand2017certicoq,
chlipala2009effective}.

Specifications come in a wide range of formats. Many, like the OpenMP
specification, are very informal, primarily consisting of english descriptions
of how different constructs should behave \cite{dagum1998openmp}. Others, like
Legion's, are written by hand in the using standard notation for logical
relations. For this paper, when we refer to a \emph{formal} specification, we
are referring to a specification defined in a formal logic, eg. Coq.

While semantics defined by hand as logical relations are a huge improvement
over natural language specifications, there is still much room for improvement.
Like type checking a program, type checking a proof exposes many "bugs" in a
proof. This is not to say, of course, that formal specifications and proofs are
infallible. On the contrary, it is quite easy to get a specification wrong.
That said, often, if a there is a mistake in a spec, the process of proving it
formally will expose that mistake, allowing for its correction.

@@

A program without a specification cannot be incorrect, it can only be
surprising \footnote{Paraphrase of J.J. Horning, 1982}. One of the most
important and challenging aspects of proving correctness is defining a formal
specification of how a program should behave. Indeed, there is currently a
large NSF effort involving many of the best minds in programming languages
dedicated to improving our ability to do just this \cite{weng2016deepspec}. The choice
of specification has far-reaching consequences in how easy it is to later prove
correctness properties.

One thing that makes Legion unique among high performance runtime systems is
that it provides a specification of its operational semantics \cite{treichler2013language}.
We take this specification of and formalize it in Coq, enabling us to prove
lemmas and theorems about it. See Figure~\ref{syntax} for the Coq
implementation of the syntax, and Figure~\ref{semantics} for semantics. All
code is available online at \texttt{https://github.com/lanl/formallegion}.

@@

The Legion semantics are defined as a big step operational semantics, which
result in non-deterministic orderings of memory reads and writes. By defining a
notion of \texttt{valid\_interleaves}, we can reason formally about what evaluations 
can happen concurrently. This should allow us to prove formally that if the
runtime makes scheduling decisions based on region permissions in a particular
way, serial semantics are preserved with respect to the resulting heap state
and return value. This is a desirable property: it means that any code using
Legion can reason, either formally or informally, about its code with respect
to a simple serial semantics, and have those semantics be preserved for
parallel execution. This is in contrast to programming models like Cilk, which
claim a weaker property: a serial execution is one of many valid executions.

@@

While so far we have discussed the formalization of the \emph{specification},
we have largely left discussion of proofs by the wayside. In this section, we
discuss the challenges of proving useful properties of real world systems, and
give some examples of basic lemmas that will be useful in proving correctness
properties of Legion. The hope is that the reader will gain some insight into
the challenges involved in proving formal properties of systems of real-world
systems.

One of the core notions of Legion is the notion of region relations. These
describe how regions are related, either by being disjoint, or by being
subregions of one another. See Figure~\ref{regionrel} for the syntax. Another
important construction are mappers, \texttt{M}, which map logical regions
\texttt{r} with a set of physical locations. We include some definitions as
well, notably the notion that a mapping respects (\texttt{resp}) a set of
region requirements iff for every subregion relation and disjoint relation, the
mapping to sets of physical locations follows the corresponding set relation.

@@

One important property when proving correctness in Legion is to show that the
relations are equivalent to their closure, with respect to what they imply
about locations in those regions.  Figure~\ref{regionrelproof} for the
statement of this proof. Intuitively, this means that if a region is a
subregion of another region, and that region is a subregion of a third, then
the first subregion is a subregion of the third. Disjointness is commutative, 
and subregions of disjoint regions are disjoint. We've left out the Coq tactic
proofs for space, as they are likely less informative to the reader than the
propositions and definitions, but they can be found online as linked above. 
Lemma \texttt{resp\_clos\_sub} proves that given that if a subregion relation
holds with respect to a mapping \texttt{M} and a set of region constraints
\textOmega{Ω} for all regions \texttt{r1, r2} then holds for the closure of that
set of region constraints under that mapping. Given \texttt{resp\_clos\_sub}, 
Lemma \texttt{resp\_clos} proves that this implies that a mapping respects
a set of region constraints if and onlyo if it respects the closure of that
set. This non-trivial property is crucial for proving that the final decisions 
made by the tasking system based on the sets of constraints preserves 
serial semantics.

@@

To do this we are writing our proofs of progress and preservation (ie. that well
typed expressions evaluate to a value or diverge and that every transition in
the operational semantics produces a well typed value given a well typed
expression) together with our implementation of the semantics of the Core
Legion language.

@@

The theorems we are using are described in Figure~\ref{preservationtheorem} and
Figure~\ref{progresstheorem} (see \texttt{https://github.com/lanl/formallegion}
for the in progress proofs).

@@

The work presented in this paper is a small fraction of a small fraction of the
amount of work required to build a complete, correct-by-construction HPC
application. In this section, we outline some of the possible paths towards
this worthy goal.

As mentioned earlier in the paper, one of the keystones for large formally
correct systems is modularity. This modularity has two purposes. On the one
hand, it enables ease of implementation. The second, analagous advantage, is
that in a similar way modularity eases the burden of proofs. By defining
specifications as abstractions, one can separate concerns of proofs. For
example, by composing a proof that a program's serial semantics comply with a
formal specification with a proof that a parallel language respects those
serial semantics, we can get a proof that a parallel program respects the
original formal semantics. In HPC, by applying this idea to parallel
programming models like Legion, we can separate the application developer from
having to prove correctness with respect to one or more increasingly complex,
heterogeneous architechures. Instead, we can provide an intermediate layer,
like Legion, that allows application developers to reason and prove correctness
to much simpler operational semantics, and then prove that those semantics are
retained even on the complex, heterogeneous architectures of todays
supercomputers.

Of course this is much easier said than done, and we address some of the challenges
faced by such an effort in the following subsections.

@@

One common theme in formalizing systems is the discovery of issues in the
corresponding informal specification. For example, when formalizing the C
language for the CompCert project, Leroy et. al. found many cases where the
informal specification was simply incomplete \cite{leroy2012compcert}. Note
that this is seperate notion from undefined behavior. Like unknown unknowns,
gaps in a specification can make it impossible to prove desired properties of a
system.  In contrast, undefined behaviour is only a problem when programs
exihibit it.

In formalizing Legion, we discovered that while there is an impressive theory
ensuring that the side effects of concurrent evaluations preserve serial
semantics with respect to memory reads and writes, there is no theory for what
expressions can be evaluated concurrently based on value dependencies. For
example, if we define the program
\begin{center}
\begin{BVerbatim}
Definition read_write x y t T r :=
  elet x t (read (id y)) 
    (write (new T r) (id x)).
\end{BVerbatim}
\end{center}
then the semantics will happily order the \texttt{read} \emph{after} the
\texttt{write}, despite the fact that the value being written, \texttt{x},
generally will need to be read before being written. This isn't catastrophic
for the semantics, as the returned value and resulting heap are equivalent, 
as desired, but it limits the ability to reason effectively about concurrency. 
In the actual implementations this is handled both dynamically with futures
and statically with dependency analysis \cite{slaughter2015regent}.

@@

A third issue is that there is no way to dynamically create and destroy
regions. In other words, if the semantics assumes static memory allocation.
Of course, this differs from the actual implementation, and is an important,
though difficult, semantics to formalize. As memory heirarchies get more 
sophisticated, the need for tools for reasoning about memory use and locality
becomes more essential.

@@

Returning to the "P" in "HPC", one important current area of research is formal
reasoning about performance \cite{mccarthy2017coq, chargueraud2015machine}. One
of the challenges for Legion, and runtime systems in general, is that the
scheduler often has a non-trivial amount of work to do, leading to large
overheads. In Legion, the combination of reasoning about permissions, memory
locality, etc., makes this overhead particularly hard to predict.

This is one oft-overlooked advantage to formal reasoning: it allows reasoning
about \emph{any} property of a system, not just correctness. For example, there
is the work by \cite{mccarthy2017coq} working towards formally verified
complexity of code.  Chlipala et. al. have also used techniques like those
described here to construct and verify hardware description, where performance
must be reasoned about down to single cycles \cite{braibant2013formal}.  This
is one area where other tasking systems, like Cilk, have a better story than
Legion. Blumofe et al. informally describe bounds on runtime overheads, which
has been realized in practice as low overheads in the actual runtime
\cite{blumofe1995cilk}.

Ideally, we'd like to make similar claims about Legion runtime overheads.
Given a formal model of the work the runtime system has to do, one can reason
about bounds on overhead. Currently, due to the complexity of the runtime
implementation, this is infeasible. Like some of the desired correctness
properties, a significant simplification of the runtime, and perhaps the
semantics, would likely be required to achieve such a goal.

@@

While we've focused on Legion, there are other parallel programming systems
that offer exciting opportunities for formalization. While some (looking at
you, OpenMP), are likely much to large and complex to formalize without
herculean effort, others have simple enough semantics that formalization is
likely within reach. One in particular that is attractive due to its simplicity
is the recent work on parallel extensions to LLVM, the Tapir project
\cite{schardl2017tapir}. Building on existing formalization efforts for LLVM
\cite{zhao2012formalizing}, extending existing semantics with the Tapir instructions for
concurrency should be viable. Again, there would be many challenges,
particularly in the area of formal reasoning about memory models, but the
approach should be viable.

As mentioned earlier, there is also the important notion of \emph{shared
semantics accross different tasking systems}. This notion of shared semantics
presents a valuable opportunity for modularization of specification. By defining
a shared semantics developers could reason more effectively, both formally and
informally, about the semantics of their code with respect to tasking layers.

@@

We hope to have convinced the reader of the importance and viability of proving
parallel programming systems in HPC correct. While we've only taken small steps
on the path to provably correct Legion, our hope is that it will inspire
support for such ventures by providing a convincing story on how to move
forward. There is ever- increasing evidence that this class of deep
specifications and certified software is an incredibly powerful tool for
proving desirable properties about our large software systems. We'd like to see
more focus on correctness in HPC, and we think deep specifications and
certified proofs are the way forward.

>>>>

\begin{abstract}
The heterogenity and massive scale of extreme scale systems are making the
traditional ad hoc implementation of programming models impractical. Formal
specification and verification of programming models is a promising approach to
ensuring correct behavior on heterogeneous systems at scales that would be
impractical for empirical verification. In this paper we formalize the semantics
of the Legion programming model which uses logical regions of memory to reason about 
locality and concurrency in a mechanized
proof assistant, Coq. We then show how this enables formal reasoning about the
model and
prove theorems about the semantics. We also discuss how to extend this work to
prove that serial semantics are preserved and how to extend the effort to
provide a complete proof of the correctness of its implementation.
\end{abstract}

@@

\section{Introduction}
As High Performance Computing (HPC) systems run increasingly complex simulations at
increasingly large scales optimization of resource usage becomes proportionally
more important in terms of both economic efficiency and scientific productivity. However the complexity of the interaction of the different
factors being optimized, such as power consumption, computational intensity and
I/O thoughput are outstripping the ability of individual researchers to understand or debug empirically.

Given the importance of importance of HPC systems to issues of both national and
economic security being able to efficiently and correctly specify, develop and
validate these systems is an important research challenge with impacts 
ranging from climate modeling to nuclear safety.

Formal specification and code proved by formal methods such as program logics
provide the strongest guarantees of correctness. Specifications describe of how a program
should behave, while proofs show that a implementations conform to
their specifications. These \emph{formal} specifications and proofs 
have been embedded in a machine-checked formal
logic \cite{bertot2013interactive, nipkow2002isabelle, pfenning1999system,
jackson1994nuprl}. This process of formalizing specifications and proofs in
machine checked logics has been shown to catch common errors in previously
informal proofs, and therefore provide higher 
assurance of correctness than alternative methods \cite{clarke1996formal,
yang2011finding}.

In addition, recent work has shown that modular, composable formal
specifications enable formal correctness proofs for large software systems that
were previously thought unattainable \cite{weng2016deepspec}. By integrating a formal
specification tightly with an implementation, one can achieve the high
levels of confidence in a codes' adherence to its
specification\cite{weng2016deepspec}. For example, the Compcert project
provides a specification for the C language, implements it as series of
composed modules which are proved correct independently before being combined
into an optimizing C compiler. The success of Compcert in removing compiler bugs has been well
documented ~\cite{leroy2012compcert}. We will discuss later how Compcert could be used in building
verifiable HPC systems.

HPC programming models are particularly attractive target for these composable
specifications and verified implementations because of their importance 
ensuring that simulation codes stay maintainable, performant, and portable at
extreme scales\cite{diaz2012survey}.
We claim in particular that formalizing programming models have wide ranging
impact across the spectrum of HPC application development --any composable correctness
proofs can be leveraged into proofs of correctness of every program using the
model rather than having to be reimplemented on a per application basis.

In this paper, we apply these composable specification and verification methods
to a recent programming model, Legion, by formalize its
semantics and type system in the Coq proof assistant. We then prove a few supporting lemmas as examples of
the kinds of things one can prove in an HPC settings. Finally, we discuss the
some of the possible paths towards a full, formally correct implementation.

This paper makes the following contributions:
\begin{itemize}
  \item An analysis of the requirements for specifying and verifying an HPC
    programming model. 
  \item A formalization of the semantics of the Legion Programming Model in the
    Coq proof assistant.
  \item A description of a path to a fully verified programming model with provable
    correctness properties using soundness and completeness proofs.
  \item An exploration of the challenges to formally specifying the Legion runtime.
\end{itemize}

@@

Legion is an HPC programming model that enables
reasoning about locality and concurrency via logical regions of memory. It has
the attractive property that it is guaranteed to preserve serial semantics for
memory operations within logical regions. This allows programmers to avoid the
burden of reasoning about the non-determinism that arises from concurrent
computation. Specifically we'll be formalizating work on from Triecher et.
al \cite{treichler2013language}. While there has been subsequent work extending
Legion's capabilities, the core semantics of the Programming Model has stayed relatively stable
\cite{slaughter2015regent, treichler2016dependent}.

By partitioning regions into \emph{subregions}, and tracking \emph{privileges},
Legion is able to ensure that certain code won't interfere or alias, and can
therefore be safely --at least with respect to those memory regions-- run in
parallel. This partitioning mechanism has been shown to be effective in enabling
applications to scale both on and off node~\cite{bauer2012legion,
treichler2014realm}. However subtle bugs still occur 
 in the implementation. One goal of this paper is to move towards
formally and systematically removing \emph{all} bugs from the programming
model, giving application developers and researcher quantifiable certainty that
their code correctly implements the semantics of the science being simulated.

Mechanized proof assistants provide one tool to reach this certainty.
Improvements to proof assistants over recent decades have made them a powerful
tool for implementing and reasoning about code. Recent research using proof
assistants has been able to verify the correctness of compilers~\cite{leroy2012compcert} and 
and operating systems~\cite{gu2011certikos}.

In particular, the Coq proof assistant has been effective for a variety of
verification tasks. Coq uses the Calculus of Inductive Constructions as its type 
theory allowing it to implement
a small dependently typed programming language and logic, capable of  
formal proofs of arbitrary mathematical objects\cite{bertot2013interactive}. In the last 10
years it has been used to both build systems via program extraction and prove
useful properties about them \cite{leroy2012compcert, anand2017certicoq,
chlipala2009effective}.

To implement a specification of a Programming Model in a proof assistant like
Coq it is customary to specify that model as a set of inference rules describing
the typing rules (the 'Statics') which describe valid terms in the model and its
operational semantics (the 'Dynamics') which describe the valid state
transitions of the model. However, most Programming Models are not specified at
this level of detail. Many, like the OpenMP
specification, are very informal, primarily consisting of English descriptions
of how different constructs should behave \cite{dagum1998openmp}. Others, like
Legion's, are written by hand in the using standard notation for logical
relations for types and operational semantics with informal proofs based on these
rules. These approaches are contrasted to \emph{formal} specifications, 
which are specifications defined in a machine checkable formal logic, eg. Coq.

The machine checkable proofs provided by these formal specifications make it
possible to encode both the natural language and logical specifications in a
formal language and verify the correctness in a manner analogous to type checking a program, exposing "bugs" in 
proofs and incompleteness in specifications. It is worth nothing, however, that
these that formal specifications and proofs implemented in proof assistant are
not infallible. Mistakes in specification can lead to incorrect assumptions and
behavior and the process of recognizing and correcting these inconsistencies is
an important part of the proving process.

@@

Experience in the specification of programs and fixing bugs in those
specification has led to the aphorism that ``program without a specification cannot be incorrect, it can only be
surprising''\footnote{Paraphrase of J.J. Horning, 1982}. Its insight is that one
of the most
important and challenging aspects of proving correctness is defining a formal
specification of program behavior. Developing correct and modular specifications
is a subject of active research including the NSF DeepSpec
project\cite{deepspecnsf} which is exploring abstraction and composition
mechanisms for program specification. Particularly important in the development
of specifications is that the assumptions stated in a 
specification have far-reaching consequences in how easy it is to later prove
correctness properties.

The Legion Programming Model provides an attractive target for formal
specification and verification because  --unique among high performance runtime systems-- Legion
provides a specification of its operational semantics \cite{treichler2013language}.
We take this specification of and formalize it in Coq, enabling us to prove
lemmas and theorems about it. Figure~\ref{syntax} provides a Coq
implementation of the syntax, and Figure~\ref{semantics} the semantics. All
code is available online at \texttt{https://github.com/lanl/formallegion}.

@@

The Legion semantics are defined as a big step operational semantics, which
result in non-deterministic orderings of memory reads and writes. By defining a
notion of \texttt{valid\_interleaves}, we can reason formally about what evaluations 
can happen concurrently. This should allow us to prove formally that if the
runtime makes scheduling decisions based on region permissions in a particular
way, serial semantics are preserved with respect to the resulting heap state
and return value. This is a desirable property: it means that any code using
Legion can reason, either formally or informally, about its code with respect
to a simple serial semantics, and have those semantics be preserved for
parallel execution. This is in contrast to programming models like Cilk, which
claim a weaker property: a serial execution is one of many valid possible executions.

@@

While so far we have discussed the formalization of the \emph{specification},
we have largely left proofs undiscussed. In this section, we
discuss the challenges of proving useful properties of real world systems, and
give some examples of basic lemmas that will be useful in proving correctness
properties of Legion. The hope is that the reader will gain some insight into
the challenges involved in proving formal properties of systems of real-world
systems.

One of the core notions of Legion is the notion of region relations. These
describe how regions are related, either by being disjoint, or by being
subregions of one another. See Figure~\ref{regionrel} for the syntax. Another
important construction are mappers, \texttt{M}, which map logical regions
\texttt{r} to a set of physical locations. We include some definitions as
well, notably the notion that a mapping respects (\texttt{resp}) a set of
region requirements iff for every subregion relation and disjoint relation, the
mapping to sets of physical locations follows the corresponding set relation.

@@

One important property to show when proving correctness of programs written the
Legion model is to show that relations are equivalent to their closure (ie. that
composed relations are equivalent to relations in isolation), especially with
respect memory locality. Figure~\ref{regionrelproof} provides a proof for this
assertion. Intuitively, this means that if a region is a
subregion of another region, and that region is a subregion of a third, then
the first subregion is a subregion of the third. Disjointness is commutative, 
and subregions of disjoint regions are disjoint. We've left out the Coq tactic
proofs for space, as they are likely less informative to the reader than the
propositions and definitions, but they can be found online as linked above. 
Lemma \texttt{resp\_clos\_sub} proves that given that if a subregion relation
holds with respect to a mapping \texttt{M} and a set of region constraints
\textOmega{Ω} for all regions \texttt{r1, r2} then holds for the closure of that
set of region constraints under that mapping. Given \texttt{resp\_clos\_sub}, 
Lemma \texttt{resp\_clos} proves that this implies that a mapping respects
a set of region constraints if and onlyo if it respects the closure of that
set. This non-trivial property is crucial for proving that the final decisions 
made by the tasking system based on the sets of constraints preserves 
serial semantics.

@@

To do this we are writing our proofs of completeness and soundness (ie. that well
typed expressions evaluate to a value or diverge and that every transition in
the operational semantics produces a well typed value given a well typed
expression) together with our implementation of the semantics of the Core
Legion language.

@@

The theorems we are using are described in Figure~\ref{soundnesstheorem} and
Figure~\ref{completenesstheorem} (see \texttt{https://github.com/lanl/formallegion}
for Coq versions of the proofs).

@@

The work presented in this paper is a small fraction of the
amount of work required to build a complete, correct-by-construction HPC
application. In this section, we outline some of the possible paths towards
this goal.

As mentioned earlier, one of the important characteristics of large formally
correct systems is modularity. This modularity has two purposes. On the one
hand, it enables ease of implementation. The second, analogous advantage, is
that modularity eases the burden of proofs. By defining
specifications as abstractions, one can separate proof concerns into proofs of
individual components or modules. For
example, by composing a proof that a program's serial semantics comply with a
formal specification with a proof that a parallel language respects those
serial semantics, it is possible derive a proof that a parallel program respects the
original formal semantics. In HPC, by applying this idea to parallel
programming models like Legion, we can separate the application developer from
having to prove correctness with respect to one or more increasingly complex,
heterogeneous architectures. Instead, we can provide an intermediate layer,
like Legion, that allows application developers to reason and prove correctness
to much simpler operational semantics, and then prove that those semantics are
retained even on the complex, heterogeneous architectures of today's
HPC Systems.

However modularly specifying HPC is not easy and is a challenging research
problem. We illustrate some of these difficulties in the following section.

@@

One common theme in formalizing systems is the discovery of issues in the
corresponding informal specification. For example, when formalizing the C
language for the CompCert project, Leroy et. al. found many cases where the
informal specification was simply incomplete \cite{leroy2012compcert}. Note
that this is separate notion from undefined behavior. Like unknown unknowns,
gaps in a specification can make it impossible to prove desired properties of a
system.  In contrast, undefined behavior is only a problem when programs
exihibit it.

In formalizing Legion, we have discovered that while there is a rich theory
ensuring that the side effects of concurrent evaluations preserve serial
semantics with respect to memory reads and writes, there is no theory for what
expressions can be evaluated concurrently based on value dependencies. For
example, if we define the program
\begin{center}
\begin{BVerbatim}
Definition read_write x y t T r :=
  elet x t (read (id y)) 
    (write (new T r) (id x)).
\end{BVerbatim}
\end{center}
then the semantics will happily order the \texttt{read} \emph{after} the
\texttt{write}, despite the fact that the value being written, \texttt{x},
generally will need to be read before being written. This inconsistency isn't catastrophic
for the semantics, the returned value and resulting heap are equivalent. 
 However it limits the ability to reason effectively about concurrency. 
In the actual implementations this is handled both dynamically with futures
and statically with dependency analysis \cite{slaughter2015regent}.

@@

A third issue is that there is no way to dynamically create and destroy
regions. In other words, if the semantics assumes static memory allocation.
Of course, this differs from the actual implementation, and is an important,
though difficult, semantics to formalize. As memory hierarchies get more 
sophisticated, the need for tools for reasoning about memory use and locality
becomes more essential.

@@

In addition to semantic correctness, one important current area of research in
formal methods is formal
reasoning about performance \cite{mccarthy2017coq, chargueraud2015machine}. One
of the challenges for Legion, and runtime systems in general, is that the
scheduler often has a non-trivial amount of work to do, leading to large
overheads. In Legion, the combination of reasoning about permissions, memory
locality, etc., makes this overhead particularly hard to predict.

This is one oft-overlooked advantage to formal reasoning: it allows reasoning
about \emph{any} property of a system, not just correctness. For example, there
is the work by \cite{mccarthy2017coq} working towards formally verified
complexity of code.  Chlipala et. al. have also used techniques like those
described here to construct and verify hardware description, where performance
must be reasoned about down to single cycles \cite{braibant2013formal}.  This
is one area where other tasking systems, like Cilk, are better understood than
Legion. Blumofe et al. informally describe bounds on runtime overheads, which
has been realized in practice as low overheads in the actual runtime
\cite{blumofe1995cilk}.

Ideally, it would be possible to make similar claims about Legion runtime overheads.
Given a formal model of the work done by the runtime system, one can reason
about bounds on overhead. Currently, due to the complexity of the runtime
implementation, this is infeasible. Like some of the desired correctness
properties, a significant simplification of the runtime, and perhaps the
semantics, would likely be required to achieve such a goal.

@@

Although the focus of this paper has been on the Legion Programming Model, there are other parallel programming systems
that offer exciting opportunities for formalization. While some, such as
 OpenMP, are likely much to large and complex to formalize without
significant research effort, others have simple enough semantics that formalization is
likely tractable.the Tapir project
\cite{schardl2017tapir} is a particularly attractive research target due to its simplicity
and the recent work on parallel extensions to LLVM. Building on existing formalization efforts for LLVM
\cite{zhao2012formalizing}, extending existing semantics with the Tapir instructions for
concurrency should be achievable in reasonable time despite challenges in
specifying its memory model.

As mentioned earlier, there is also the important concept of \emph{shared
semantics accross different tasking systems}. This notion of shared semantics
presents a valuable opportunity for modularization of specification. By defining
a shared semantics developers could reason more effectively, both formally and
informally, about the semantics of their code with respect to tasking layers.

@@

In this paper, we have proposed a path to formally specifying HPC programming
models using mechanized proof assistants. To evaluate this proposal we have given a formal
semantics of the Legion Programming Model and proved its correctness regarding
memory interleavings between relations. We have also described mechanisms to
prove the soundness and completeness of our approach as well as the potential
difficulties in dealing with potential gaps in specification. Finally we
concluded by proposing the Tapir system as a viable next step in the
specification of HPC programming models.
